Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-02-27 resex.tex
Revised Section IV on experimental setup.

\begin{figure*}
  \includegraphics[width=\textwidth]{bunchfilling_measured.png}	
  \caption{Example of bunch-by-bunch excitation detected by the
    real-time transverse activity monitor during the 2017 experiment,
    when the beam was excited every $8^\mathrm{th}$~turn.}
  \label{fig:fill_meas} 
\end{figure*}

@@

\subsection{Excitations with the transverse damper and filling schemes}
\label{sec:adt}

The primary function of the transverse feedback system in LHC, also
referenced to as ADT, is to provide injection oscillation damping and
actively damp the coupled bunch instabilities driven by the machine
impedance. The main building blocks of the ADT system are stipline
pickups at the position Q7 and Q9 at Point~4 of the LHC, connected to
the beam position measurement modules at the surface, the digital
signal processing modules (mDSPU) and a set of tetrode power
amplifiers feeding electrostatic kickers in the RF zone of the
LHC~\cite{adt_sum_2008,adt_sum_2011}. Though originally designed to
damp the undesired transverse activities only, the damper is being
routinely used for sophisticated beam excitations as well thanks to
the state of the art digital signal processing hardware. These include
among others abort and injection gap cleaning, single bunch excitation
for tune and linear coupling measurements and special modes of
operation for dedicated experiments at the LHC.

The transverse feedback is in general active during the whole LHC
cycle, therefore it is critical that the noise introduced by the
system does not contribute to any emittance growth. The typical
machine cycle requires a short damping time of ~10-20 turns (high
feedback gain) for injection oscillation damping, which is increased
during the ramp and collisions to ~50-100 turns (lower feedback
gain). It has been demonstrated that noise introduced by the
transverse feedback system itself is sufficiently low not to introduce
any measurable emittance growth \cite{adt_noise_emit_2017}. All
excitation signals referenced to in this paper are digitally
synthesized in the damper's digital signal processing units and
therefore can be considered as ``noise-free''. This point is important
as then the effect on the beam can be attributed entirely to the
applied excitation and any effect due to unwanted residual noise can
be neglected.

The resonant excitation experiment in 2017 involved simultaneous
measurements on three groups of 72 bunches with a dedicated excitation
pattern and transverse feedback configuration on each subset of 6
bunches. In 2016 a similar configuration was used, but with only 48
bunches in total and subsets of 4 bunches. Both schemes are
illustrated in Fig.~\ref{fig:fill}. In order to reduce the error of
the measurements, the observables are averaged over each subset of
bunches experiencing the same excitation amplitude, yielding in total
5~different amplitudes $n\Delta A$ with $n=1,\cdots, 5$ plus the
references bunch without excitation for each transverse feedback
configuration and excitation plane. Furthermore, the experiments were
usually split in 3 different parts:
\begin{description}
	\item[Part 1] No excitation is applied in order to allow the distribution to reach an equilibrium state after injection.
	\item[Part 2] The excitation is applied with a first maximum excitation amplitude of $A_{\mathrm{max,1}}=5\Delta A_1$.
	\item[Part 3] The maximum excitation amplitude is further increased to $A_{\mathrm{max,1}} < A_{\mathrm{max,2}}=5\Delta A_2$.
\end{description}
Each part lasted approximately 10-15~minutes, which is considered to
be a long enough time window to allow the beam distribution to reach
its new equilibrium state. In the plots of the experimental results
shown later in the paper, the maximum excitation amplitude is
indicated with black horizontal arrows, while the average of each
subset of bunches with the same excitation amplitude $n\Delta A$ is
distinguished with different colors.

The excitation in the horizontal and the vertical plane is generated
by a different set of signal processing hardware, however during this
experiment the signals were synchronized at a turn level. This means
the bunches of the third group (of 72 bunches) experienced the kick in
both horizontal and vertical plane during the same turn.

The ADT kicker deflection angle is calculated from the kicker geometry
and the excitation voltage on the deflection plates. The excitation
voltage used in the experiments is a complicated waveform, generated
by a very complex hardware (digital signal processor, chain of
transmission lines, low and high power amplifiers). The peak
excitation voltage (and subsequently the kick) seen by the beam can be
only indirectly measured using the probes on each kicker. The measured
values are therefore believed to be known with an error margin of
10-15\% for the experiments in 2017 and 50\% in 2016. The estimate for
2017 is based on only a single measurement and in the future and with
more time available the determination of the error would definitely
have to be improved. The error in 2016 is estimated from the
transverse damper operational parameters. Due to lack of the dedicated
machine time no precision calibration was performed. The maximum kick
strength reachable with the ADT at injection without risking
saturation is 96~nrad. As an example Fig.~\ref{fig:fill_meas} shows
the oscillation amplitude of each bunch in both H and V planes
captured by the LHC real time transverse activity monitor during the
2017 experiments and an excitation with $8^{\mathrm{th}}$ turn
pulsing. The measured activity is well in line with the expected
excitation pattern.

>>>>

\begin{figure*}
  \includegraphics[width=\textwidth]{bunchfilling_measured.png}	
  \caption{Example of bunch-by-bunch beam centroid motion for Beam~1
    in the horizontal (B1H, left) and vertical (B1V, right) planes,
    detected by the real-time transverse activity monitor during the
    2017 experiment, when the beam was excited every
    $8^\mathrm{th}$~turn.}
  \label{fig:fill_meas} 
\end{figure*}

@@

\subsection{Excitations with the transverse damper and bunch filling schemes}
\label{sec:adt}

The primary function of the LHC transverse damping and feedback
system, also known as ADT, is to mitigate injection oscillations and
to actively damp the coupled-bunch instabilities driven by machine
impedance~\cite{adt_sum_2008, adt_sum_2011}. The main building blocks
of the system are the following: strip-line pickups at positions Q7
and Q9 near Interaction Point~4 (IP4) of the LHC, which are connected
to the beam position measurement modules at the surface; the digital
signal processing modules (mDSPU); and a set of tetrode power
amplifiers feeding electrostatic kickers in the same radio-frequency
sector of the LHC (IP4).

Because of its flexibility and state-of-the-art hardware, the system
is being routinely used for sophisticated beam excitations. These
include abort- and injection-gap cleaning, excitation of individual
bunches for tune and linear-coupling measurements, and other special
modes of operation for dedicated experiments in the LHC.

The transverse feedback is in general active during all phases of LHC
operation. The typical machine cycle requires a short damping time of
10--20 turns (high feedback gain) for injection oscillation
damping. During the acceleration ramp and during collisions, the
damping time is increased to 50--100 turns (lower feedback gain).

Because the damper is always active in operations, it is critical that
the noise introduced by the system does not cause any measurable
emittance growth, and this fact was verified
experimentally~\cite{adt_noise_emit_2017}. All excitation signals
described in this paper were digitally synthesized in the ADT's
digital signal processing units and are therefore assumed to be
`noise-free.' The observed effects on the beam are attributed to the
applied excitations and the effects of unwanted residual noise are
assumed to be negligible.

The resonant excitation experiment in~2017 involved simultaneous
measurements on 3~groups of 72~bunches, with dedicated excitation
patterns and transverse feedback configurations (i.e., damper active
or damper off) on each subset of 6~bunches. In~2016, a similar
configuration was used, with 48~bunches in total and subsets of
4~bunches. Both schemes are illustrated in Fig.~\ref{fig:fill}. A
total of 5 different amplitudes could be applied simultaneously to
different subsets of bunches, denoted as $n \, \Delta A$, with
$n = 1, \ldots, 5$. In addition, there were reference bunches without
excitation for each transverse feedback configuration and excitation
plane. Observables (losses, emittances, etc.) were averaged over each
subset of bunches experiencing the same excitation and damping
conditions.

In~2017, the excitations in the horizontal and vertical planes were
generated by a different set of signal-processing devices, but
synchronized at a turn-by-turn level. Therefore, the bunches of the
third group of 72 bunches (see Fig.~\ref{fig:fill}) were affected by
the kicks in both horizontal and vertical planes during the same turn.

The experiments were usually split in time into 3 different periods:
%
\begin{description}
\item[Period 1] No excitation was applied, to allow the beam to reach
  an equilibrium state after injection.
\item[Period 2] The excitation was applied with a first maximum
  excitation amplitude of $A_{\mathrm{max,1}} = 5 \, \Delta A_1$.
\item[Period 3] The maximum excitation amplitude was further increased
  to $A_{\mathrm{max,2}} = 5 \, \Delta A_2 > A_{\mathrm{max,1}}$.
\end{description}
%
Each period lasted approximately 10--15~minutes, which was considered
long enough to allow the beam to reach its new equilibrium state. In
the discussion of the experimental results (Section~\ref{sec:simex}
below), the 3~periods are labeled according to the maximum excitation
amplitude~$A_\mathrm{max}$, and subsets of bunches with the same
excitation amplitude $n \, \Delta A$ ($n = 1, \ldots, 5$) are grouped
by color.

The proton deflection angle generated by the ADT kicker was calculated
from the kicker geometry and from the excitation voltage on the
deflection plates. The voltage depends on a complex chain of hardware
(digital signal processor, transmission lines, low- and high-power
amplifiers, etc.). In~2016, the kick was estimated from the
operational system parameters and it was assigned an uncertainty of
50\%. In~2017, the excitation voltage could be indirectly measured
once, using the probes mounted on the kickers, with an estimated
uncertainty of 10--15\%. A precise in-situ calibration was not
possible due to the limited machine availability for dedicated
studies.

The maximum ADT kick strength that could be obtained at injection
without risking saturation was approximately 100~nrad. For the
experiments, a maximum nominal kick strength of 96~nrad was chosen.

Figure~\ref{fig:fill_meas} shows an example of the oscillation
amplitude of each bunch centroid in both horizontal and vertical
planes, captured by the LHC real-time transverse activity monitor
during the 2017 experiments, when the beam was excited every
$8^{\mathrm{th}}$~turn. The measured activity shows the typical
amplitude of bunch centroid excursions and reflects the expected
excitation pattern.
